---
title: "MATH70076: Data Science - Formative Assessment"
subtitle: 'MSc in Statistics 2026/27, Imperial College London'
author: "YOUR CID"
format:
  html:
    toc: true
    highlight: tango
    self-contained: true
  pdf: default
format-links: false
bibliography: example.bib 
---

**Deadline: Monday 12 October 2026 at 13:00.**

*For this formative assessment you should submit two files via the Imperial College VLE on Canvas by the deadline stated above. Your files should be named as follows:*

- `YOURCID-MATH70076-formative-assessment.pdf`: your rendered report,
- `YOURCID-MATH70076-formative-assessment.zip`: a zip file containing the relevant source code to generate your report.

*All submitted materials should be clearly presented and be understandable as stand-alone documents.*

*Please note that large files can take quite some time to upload. Ensure that you upload each document to the correct part of the learning space in a timely manner.*

*This mock coursework is a formative assessment and so will not contribute to your overall grade for the module. It is designed for you to practice the skills you have learned in the first workshop and to get feedback on your reflective writing style. This activity is expected to take approximately 6 hours of individual effort. Assessment criteria are given in the "set yourself up for success" boxes. Satisfying 18 or more out of these 23 criteria corresponds to a notional "pass" grade.*

*In submitting this mock assessment you certify that it is entirely your own work, apart from where otherwise acknowledged, and includes no plagiarism. Note that software tools are used as part of plagiarism detection.*

------------------------------------------------------------------------

## Background

### Generalised Pareto Distribution

The Generalized Pareto Distribution (GPD) is a flexible family of continuous probability distributions that arises naturally in extreme value theory, particularly for modelling the distribution of excesses over a threshold. It is parametrised by a shape parameter $\xi \in \mathbb{R}$, a scale parameter $\sigma > 0$, and a location parameter $u \in \mathbb{R}$. Its cumulative distribution function (CDF) is given by

$$
F(x;\,\sigma,\xi, u) = 
\begin{cases}
1 - \left(1 + \dfrac{\xi (x-u)}{\sigma}\right)_+^{-1/\xi}, & \xi \neq 0, \, x \geq u; \; \\[1.2em]
1 - \exp\!\left(-\dfrac{x-u}{\sigma}\right), & \xi = 0, \, x \geq u;
\end{cases}
$$ {#eq-gpd-cdf}

where $x_+ = \max(x,0)$ and its probability density function (PDF) is

$$
f(x;\sigma,\xi,u) =
\begin{cases}
\dfrac{1}{\sigma}\left(1 + \dfrac{\xi (x-u)}{\sigma}\right)_+^{-1/\xi - 1}, & \xi \neq 0, \\[1.2em]
\dfrac{1}{\sigma}\exp\!\left(-\dfrac{x-u}{\sigma}\right), & \xi = 0,
\end{cases}
$$ {#eq-gpd-pdf}

defined on the same support as the CDF. The GPD encompasses a variety of tail behaviours:

- when $\xi > 0$ the GPD has heavy, slowly decaying tails,
- in limiting case when $\xi \rightarrow 0$ the GPD reduces to an exponential distribution;
- when $\xi < 0$ the GPD has light, quickly decaying tails with a finite upper endpoint of $x^+ = u - \sigma / \xi$.

### Probability Integral Transform

The probability integral transform states that if a random variable $X$ has a continuous cumulative distribution function $F_X(x)$, then the transformed variable $A = F_X(X)$ follows a uniform distribution on $[0,1]$. Conversely, if $F$ is an invertible function then $Y = F^{-1}_X(A)$ has the same distribution as $X$.

## Questions

### Question 1

Derive an expression for the inverse cumulative distribution function (also known as the quantile function) $F^{-1}_X: [0,1] \rightarrow [u, x^+]$ of $X \sim \text{GPD}(u, \sigma, \xi)$. Your answer should refer to at least one equation given in the background material.

::: callout-tip
## Set yourself up for success

Does your answer contain:

- a few sentences of text describing your approach to the problem;
- a reference to at least one equation from the background section;
- a correctly formatted LaTeX equation;
- a valid approach to the problem and correct expression.
:::

### Answer 1

The problem requires to derive an inverse cumulative distribution function for the Generalized Pareto Distribution. In order to derive one we must first define a new variable $q \in [0,1]$ such that $F_X(x) = q$ and solve for $x$ in terms of $q$, which would give us $F^{-1}_X$. Since @eq-gpd-cdf has tow cases, we must derive $F^{-1}_X$ for each of them, starting with the simpler case when $\xi = 0$.

#### Case $\xi = 0$:

In this case we have that $F_X(x) = q = 1 - \exp\!\left(-\dfrac{x-u}{\sigma}\right)$ with $x \geq u$, so solving for $x$ we get

$$F^{-1}_X(q) = u - \sigma \ln(1 - q)$$

#### Case $\xi \neq 0$:

In this case we have that $F_X(x) = 1 - \left(1 + \dfrac{\xi (x-u)}{\sigma}\right)_+^{-1/\xi}$ with $x \geq u$. This case has an expression of the form $x_+ = \max(x,0)$ meaning it can also be divided in two sub cases:

$$
\left(1 + \dfrac{\xi (x-u)}{\sigma}\right) \gt 0
$$ {#eq-max-expr-gt-zero} and $$
\left(1 + \dfrac{\xi (x-u)}{\sigma}\right) \leq 0
$$ {#eq-max-expr-leq-zero}

Since we have that $x \geq u$ and $\sigma > 0$ we know that if $\xi >0$ then @eq-max-expr-gt-zero holds, so solving for $x$ gives us

$$
F^{-1}_X(q) = \dfrac{\sigma}{\xi} \left(\dfrac{1}{(1-q)^\xi} - 1 \right) + u
$$ {#eq-qdist-max-expr-gt-zero}

If $\xi \lt 0$ then we have to find the values of $x$ for which @eq-max-expr-leq-zero holds. By solving the inequality we get that @eq-max-expr-leq-zero holds when $x \geq u + \dfrac{\sigma}{|\xi|}$, which would give us that $F_X(x) = 1$.

Otherwise, @eq-max-expr-gt-zero is true when $u \leq x \lt  u + \dfrac{\sigma}{|\xi|}$ so we are back with the solution @eq-qdist-max-expr-gt-zero.

Now using $F_X(x)$ to get the conditions in terms of $q$ we get that $F_X(u) = 0$ for all cases and $F_X(x) = 1$ when $\xi \neq 0, \, x \geq u + \dfrac{\sigma}{|\xi|}$

Finally we put all cases together and get

$$
F^{-1}_X(q) = 
\begin{cases}
u - \sigma \ln(1 - q), & \xi = 0, \, 0 \leq q \lt 1; 
\\[1.2em]
\dfrac{\sigma}{\xi} \left(\dfrac{1}{(1-q)^\xi} - 1 \right) + u, & \xi \neq 0, \, 0 \leq q \lt 1;
\\[1.2em]
u + \dfrac{\sigma}{|\xi|}, & \xi \lt 0, \, q = 1;
\end{cases}
$$

### Question 2

*For this question you should display the R code you use to define and document `qgpd()` within the main text of your report.*

(a) Write and document your own function `qgpd()` to calculate quantiles of a given generalised Pareto distribution. Your function should have inputs and behaviour similar to the built-in R functions such as `qnorm()` and `qunif()` and check that inputs are in the correct format.

(b) Suppose a random variable $X$ follows an generalised Pareto distribution with threshold parameter $u=1.5$, scale parameter $\sigma = 2$ and shape parameter $\xi = -0.4$. Use your function to find quantiles $x_p$ for $p = 0.5, 0.75, 0.99$ (i.e. for each $p$ find the value for which $\Pr(X < x_p) = p$).

::: callout-tip
## Set yourself up for success

Does your answer:

- contain a valid R function definition;
- document the expected inputs, outputs and behaviours of that function;
- check the validity of inputs;
- handle any edge-cases in an appropriate way;
- return the correct quantile values?
:::

```{r}
# YOUR CODE GOES HERE
```

### Question 3

*For this question, all R code should be displayed only within the appendix, not in the main report.*

The file `gpd_samples.csv` contains six sets of random variates generated from different generalised Pareto distributions. The details of the generalised Pareto distributions used are summarised in `gpd_parameters.csv`. Unfortunately some of the parameter sets were recorded incorrectly.

Within a single figure, construct a series of quantile-quantile plots to identify which datasets are inconsistent with their stated distributions. You should both justify your conclusions and describe your level of confidence in your findings.

::: callout-tip
## Set yourself up for success

Does your solution contain:

- at least one quantile-quantile plot;
- six qq-plots in a single figure;
- use of loops, vectorisation, or function definitions to avoid repetitive code.
- figures with clear text, useful captions and appropriate visual mapping of data;
- a few paragraphs describing and justifying your findings and referencing the figure;
:::

<!-- YOUR ANSWER GOES HERE -->

### Question 4

*For this question, all R code should be displayed only within the appendix, not in the main report.*

A hydrologist is interested in understanding the river flow at a location that is historically prone to flooding. There is a river flow gauge nearby which measures river flow in units of cubic meters per second ($\text{m}^3/\text{s}$ or cumecs). Based on her knowledge of other rivers, the hydrologist proposes that for this river gauge:

- the distribution of river flow values is constant over time
- for river flows exceeding 75 cumecs, it is appropriate to model these data as independent and identically distributed GPD($\sigma =29.7$, $\xi=0.62$, $u = 0$).

Use `riverflow_2015-2024.csv` to conduct an exploratory investigation of whether these proposals are valid for the dataset provided. Summarise your findings in 250-350 words, supporting these with a collection of 4 visualisations/figures.

::: callout-tip
## Set yourself up for success

Does your answer contain:

- 400-500 words of text clearly describing your choice of visualisations, their interpretation and your conclusions about the validity of each assumption,
- A series of 4 visualisations / figures,
- A varied and appropriate choice of figures to investigate each of the hydrologist's proposals,
- Figures with clear text, useful captions and appropriate visual mapping of data;
- At least one reference to each visualisation within the main text.
:::

<!-- YOUR INVESTIGATION GOES HERE -->

::: callout-tip
## Set yourself up for success

- Does your document render without any formatting issues?
:::

## Question 5 Reflective Summary

Describe, evidence and reflect on the process of constructing the visualisations in your answer to Question 4.

Your response should be 300-400 words and might, for example, comment on:

- why you chose these forms of visualisation and what alternatives you considered;
- a challenge you faced or something that you learned while making the visualisations;
- how you might approach this task differently with more time or for a different intended audience.

::: callout-tip
## Set yourself up for success

Does your response:

- describe how you engaged with the visualisation activity;
- give specific evidence of that engagement;
- critically reflection on that engagement?
:::

*End of Assessment.*

# Code Appendix {#sec-code-appendix}

```{r ref.label=knitr::all_labels()}
#| echo: true
#| eval: false
#| code-fold: true
```
